Higher-Order unification for free!

### 2024-04-11 paper/main.tex
wip

\noindent
The right hand side of a link, the problematic term, can occur under binders.
To accommodate this situation the compiler wraps \elpiIn{baselink} using
the \elpiIn{inctx} container (see, \ref{data:inctx}).

@@

In the example above, we see that $\foUnifPb_1$ instantiates $X$ so that
$\foUnifPb_2$ can be solved in \llambda.
% \noteE[inline]{it is even a ground term, there is no unification left to perform actually}
% \noteD[inline]{i don't understand the note}
On the other hand, we see that, 
\Ue can't solve the compiled problems \hoUnifPb. In
fact, the resolution of $\hoUnifPb_1$ gives the substitution $\sigma = \{ A
\mapsto \lambda x. B\}$, but the dereferencing of $\hoUnifPb_2$ gives the 
non-unifiable problem $(\lambda x. B) \appsep a \nUe a$.

@@

\noteE[inline]{Dire che maybe eta fa il detect anche su 
termini che non sono il llambda, oppure dirlo in section of maybeeta + dare un
esempio?}

@@

\newcommand{\progBetaLL}{\emph{progress-beta-\llambda}\xspace}
\newcommand{\progBetaRH}{\emph{progress-beta-rigid-\rhs}\xspace}
\newcommand{\progBetaDedup}{\emph{progress-beta-dedup}\xspace}
\newcommand{\progBetaLH}{\emph{progress-rigid-\lhs}\xspace}

The activation of a \linkbeta is performed when its \rhs falls under \llambda
under a given substitution.

\begin{definition}[\progBetaLL]
  Given a substitution $\sigma$ and a \linkbeta \linkbetaM{\Gamma}{T}{\rhsBeta}
  such that $\sigma t_1$ is a name, say $t$, and $t\notin s_1\dots s_n$. If $m =
  0$, then the \linkbeta is removed and \lhs is unified with \rhsBetaHead.
  If $m > 0$, then the \linkbeta is replaced by a refined version
  $\linkbetaM{\Gamma}{T}{Y_{s_1\dots s_n,t}\appsep t_2\dots t_m}$ with reduced
  list of arguments and $Y$ being a fresh variable. Moreover, the new link
  \linketaM{\Gamma}{X_{s_1\dots s_n}}{\lambda x.Y_{s_1\dots s_n,x}} is added to 
  \linkStore.

  % A link \linkbetaM{\Gamma}{T}{\rhsBeta} is removed from \linkStore, if given
  % the substitution $\sigma$, $\sigma t_1$ is a name, say $t$, such that, $t
  % \notin s_1\dots s_n$. In this case, let $Y$ a fresh variable, then
  % \linketaM{\Gamma}{X_{s_1\dots s_n}}{\lambda x.Y_{s_1\dots s_n,x}} is added to
  % \linkStore, and 1) if $m = 1$ then $Y_{s_1\dots s_n, t}$ is unified with $T$,
  % 2) otherwise, the refined \linkbeta, \linkbetaM{\Gamma}{T}{M_{s_1\dots
  % s_n,t}\appsep t_2\dots t_m}, is added to \linkStore.
  \label{def:progBetaLL}
  % % \noteE[inline]{non è chiaro. la sostituzione immagino sia la corrente, non se ne esiste una. Poi non è chiaro se otherwise si riferische a in this case. Usa un enumerate}
  % %\noteE[inline]{Is it clearer?}
  % \noteE[inline]{l1 or t1? Forse è più chiaro dire che il ilnk beta viene raffinato con link beta cone meno argomenti, Se poi il numero di argomenti extra è 0 il tutto è in llambda e quindi viene buttato via. In questo modo la terminazione è ancora più chiara perchè si vede già che prima decresce la lista di argomenti e poi il numero di beta}
  \noteD{L'ho riformulato}
\end{definition}

\begin{definition}[\progBetaRH]
  A link \linkbetaM{\Gamma}{X}{\rhsBeta} is removed from
  \linkStore if \rhsBetaHead is instantiated to a term $t$ and the
  $\beta$-reduced term $t'$ obtained from the application of $t$ to
  $l_1\dots l_m$ is in \llambda. Moreover, $X$ is unified to $t$.
  \label{def:progBetaRH}
\end{definition}

% Example of this `X a = Y a'
\begin{definition}[\progBetaDedup]
  Given a \linkbeta $l_1$ and second link $l_2 \in\linkStore$, such that they
  share the same \lhs. The two \rhs are unified and a $l_2$ is
  removed from \linkStore.
  \label{def:progBetaDedup}
\end{definition}

@@

  % \Cref{def:progBetaRH} makes \elpiIn{progress} terminates, since it makes a
  % \linkbeta disappear from \linkStore. On the other hand, \cref{def:progBetaLL},
  % creates each time a new \linketa and replace the old \linkbeta with a new one.
  % This progression can however triggered at most $m$ times, since each new
  % \linkbeta as a smaller list of applied variables. At the $m^{th}$ progression,
  % the \linkbeta is removed from \linkStore which is now filled by $m$ new
  % \linketa. By \cref{lemma:prog-eta-terminates}, we know that \elpiIn{progress}
  % terminates if \linkStore is made by only \linketa and therefore
  % \elpiIn{progress} still terminates.
  %\noteE[inline]{is it ok?}
  \noteE[inline]{funziona. per essere più precisi io parlerei di ordine lessicografico (tipico ordine ben fondato usato per dimostrare terminazione). Nl nostro caso è la tripla (argomenti extra dei beta, numero di beta, numero di eta).}
  \noteD{da finire}
\end{proof}

>>>>

\noindent
The right hand side of a link, the problematic term, can occur under binders.
To accommodate this situation the compiler wraps \elpiIn{baselink} using
the \elpiIn{inctx} container (see rule~\ref{data:inctx}).

@@

In the example above, we see that $\foUnifPb_1$ instantiates $X$ so that
$\foUnifPb_2$ can be solved in \llambda.
% \noteE[inline]{it is even a ground term, there is no unification left to perform actually}
% \noteD[inline]{i don't understand the note}
\noteD{Cambiare esempio riprendendo quello di enrico}
On the other hand, we see that, 
\Ue can't solve the compiled problems \hoUnifPb. In
fact, the resolution of $\hoUnifPb_1$ gives the substitution $\sigma = \{ A
\mapsto \lambda x. B\}$, but the dereferencing of $\hoUnifPb_2$ gives the 
non-unifiable problem $(\lambda x. B) \appsep a \nUe a$.

@@

@@

\newcommand{\progBetaLL}{\emph{progress-beta-\llambda}\xspace}
\newcommand{\progBetaRH}{\emph{progress-beta-rigid-\rhs}\xspace}
\newcommand{\progBetaDedup}{\emph{progress-beta-dedup}\xspace}
\newcommand{\progBetaLH}{\emph{progress-beta-rigid-\lhs}\xspace}

We give $4$ different definitions of \linkbeta progression for a given \linkbeta
\linkbetaM{\Gamma}{T}{\rhsBeta}, say $l$.

\begin{definition}[\progBetaLL]
  Given a substitution $\sigma$, where $\sigma t_1$ is a name, say $t$, and
  $t\notin s_1\dots s_n$. If $m = 0$, then $l$ is removed and \lhs is
  unified with \rhsBetaHead. If $m > 0$, then $l$ is replaced by a
  refined version $\linkbetaM{\Gamma}{T}{Y_{s_1\dots s_n,t}\appsep t_2\dots
  t_m}$ with reduced list of arguments and $Y$ being a fresh variable. Moreover,
  the new link \linketaM{\Gamma}{X_{s_1\dots s_n}}{\lambda x.Y_{s_1\dots s_n,x}}
  is added to \linkStore.

  % A link \linkbetaM{\Gamma}{T}{\rhsBeta} is removed from \linkStore, if given
  % the substitution $\sigma$, $\sigma t_1$ is a name, say $t$, such that, $t
  % \notin s_1\dots s_n$. In this case, let $Y$ a fresh variable, then
  % \linketaM{\Gamma}{X_{s_1\dots s_n}}{\lambda x.Y_{s_1\dots s_n,x}} is added to
  % \linkStore, and 1) if $m = 1$ then $Y_{s_1\dots s_n, t}$ is unified with $T$,
  % 2) otherwise, the refined \linkbeta, \linkbetaM{\Gamma}{T}{M_{s_1\dots
  % s_n,t}\appsep t_2\dots t_m}, is added to \linkStore.
  \label{def:progBetaLL}
  % % \noteE[inline]{non è chiaro. la sostituzione immagino sia la corrente, non se ne esiste una. Poi non è chiaro se otherwise si riferische a in this case. Usa un enumerate}
  % %\noteE[inline]{Is it clearer?}
  % \noteE[inline]{l1 or t1? Forse è più chiaro dire che il ilnk beta viene raffinato con link beta cone meno argomenti, Se poi il numero di argomenti extra è 0 il tutto è in llambda e quindi viene buttato via. In questo modo la terminazione è ancora più chiara perchè si vede già che prima decresce la lista di argomenti e poi il numero di beta}
  % \noteD{L'ho riformulato}
\end{definition}

\begin{definition}[\progBetaRH]
  $l$ is removed from
  \linkStore if \rhsBetaHead is instantiated to a term $t$ and the
  $\beta$-reduced term $t'$ obtained from the application of $t$ to
  $l_1\dots l_m$ is in \llambda. Moreover, $X$ is unified with $t$.
  \label{def:progBetaRH}
\end{definition}

% Example of this `X a = Y a'
\begin{definition}[\progBetaDedup]
  Given a \linkbeta $l_1$ and second link $l_2 \in\linkStore$, such that they
  share the same \lhs. The two \rhs are unified and $l_2$ is
  removed from \linkStore.
  \label{def:progBetaDedup}
\end{definition}

@@

  % \Cref{def:progBetaRH} makes \elpiIn{progress} terminates, since it makes a
  % \linkbeta disappear from \linkStore. On the other hand, \cref{def:progBetaLL},
  % creates each time a new \linketa and replace the old \linkbeta with a new one.
  % This progression can however triggered at most $m$ times, since each new
  % \linkbeta as a smaller list of applied variables. At the $m^{th}$ progression,
  % the \linkbeta is removed from \linkStore which is now filled by $m$ new
  % \linketa. By \cref{lemma:prog-eta-terminates}, we know that \elpiIn{progress}
  % terminates if \linkStore is made by only \linketa and therefore
  % \elpiIn{progress} still terminates.
  %\noteE[inline]{is it ok?}
  \noteE[inline]{funziona. per essere più precisi io parlerei di ordine lessicografico (tipico ordine ben fondato usato per dimostrare terminazione). Nl nostro caso è la tripla (argomenti extra dei beta, numero di beta, numero di eta).}
\end{proof}
